---
title: "126 Data Project, Step 4"
date: "Sam Ream, Valeria Lopez, Skyler Yee"
output:
  pdf_document:
    latex_engine: xelatex
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
# knit options
knitr::opts_chunk$set(echo = F,
                      results = 'markup',
                      fig.width = 4,
                      fig.height = 3,
                      fig.align = 'center',
                      message = F,
                      warning = F)

# packages
library(tidyverse)
library(faraway)
library(RSQLite)
library(skimr)
library(GGally)
library(tidymodels)
library(leaps)
library(glmnet)
```
```{r, echo=FALSE}
ball <- dbConnect(drv=RSQLite::SQLite(), dbname="../../data/database.sqlite")
```
```{r, echo =FALSE}
batting <- dbGetQuery(ball, "
                            SELECT
                            sum(ab) AS AT_BAT,
                            player_id,
                            sum(r) AS RUNS,
                            sum(hr) AS HOME_RUNS, 
                            sum(triple) AS TRIPLE,
                            sum(double) AS DOUBLE,
                            (sum(h) -  sum(hr) - sum(triple) - sum(double)) AS SINGLES,
                            sum(bb) AS WALKS,
                            sum(ibb) AS INT_WALKS,
                            sum(sb) AS STOLEN_BASES,
                            sum(hbp) AS HIT_BY_PITCH
                            
                            FROM
                            batting
                            where
                            year > 2000
                            group by
                            player_id
                            
                   ")
sumbat <- batting
batting <- subset(batting,batting[,1]>100)


for (x in 3:11) 
{
  batting[,x] <- batting[,x] / 1
}

set.seed(5)
batting <- sample_n(batting, 500, replace = FALSE)

players <- dbGetQuery(ball, "
                      
                      SELECT 
                      player_id,
                      (weight / POWER(height, 2)) *703 AS BMI,
                      bats as HAND
                      FROM
                      player
                      
                      ")




players <- subset(players, players[,2]>0)



for ( x in 1:17918)
{
  if(players[x,2] <= 18.5) {players[x,2] <- "U"} 
  else if(players[x,2] <= 24.9) {players[x,2] <- "H"}
  else if(players[x,2] <= 29.9) {players[x,2] <- "O"}
  else {players[x,2] <- "B"}
}

batting <-  merge(batting, players, by="player_id" )
```



```{r eval = FALSE}
#Final Model
stat_model <- lm(RUNS~HOME_RUNS + SINGLES + WALKS + STOLEN_BASES, data = batting)
summary(stat_model)
```


```{r}
fmod <- lm(RUNS ~ HOME_RUNS + TRIPLE + DOUBLE + SINGLES + WALKS + INT_WALKS + STOLEN_BASES + HIT_BY_PITCH, data=batting)
summary(fmod)
```
intro: 

Using the "History of Baseball" data set, we analyzed how our predictors (singles, doubles, triples, home runs, walks, intentional walks, hit by pitches, stolen bases, BMI, and batting hand) affected the runs scores by individual players. We sampled player statistics randomly from games played between 2000-2015, which allowed us to get an accurate representation of the population of all players who played between 2000 and 2015. Using both Ridge Regression and LASSO, we shrunk the size of some predictors to obtain estimates with smaller variance for higher precision.

#Conclusion 
The data was what we had anticipated because the models had high R squared values, which indicates a large quantity of the variability can be explained by the regression. Using Ridge Regression, which aims to minimize SSE, we saw an R-squared value of 0.9914. Using LASSO regression, which shrinks the less important coefficients to zero, we got an R-squared value of 0.9936.

**Innovation Step: Principal Components**

We chose Principal Components Analysis for our innovation because 

```{r}
install.packages("corrr")
library('corrr')
install.packages("ggcorrplot")
library(ggcorrplot)
install.packages("FactoMineR")
library("FactoMineR")
install.packages("factoextra")
library("factoextra")
```
```{r}
head(batting)
numerical_data <- batting[,4:11]
head(numerical_data)
data_normalized <- scale(numerical_data)
head(data_normalized)
data.pca <- princomp(data_normalized)
summary(data.pca)
data.pca$loadings[, 1:2]
```
```{r}
# visualization
fviz_eig(data.pca, addlabels=TRUE)
fviz_pca_var(data.pca, col.var="black")
fviz_cos2(data.pca, choice="var", axes=1:2)
fviz_pca_var(data.pca, col.var = "cos2",
            gradient.cols = c("black", "orange", "green"),
            repel = TRUE)

# Doubles contributes the most; home runs, walks, singles, triples, and stolen bases also contribute highly
# home runs, walks, doubles, singles positively correlated
# triples, singles, stolen bases negatively correlated
# all have similar magnitude (distance from origin)

# chose home runs, walks, doubles 
```


