---
title: "126 Data Project, Step 2"
date: "Sam Ream, Valeria Lopez, Skyler Yee"
output:
  pdf_document:
    latex_engine: xelatex
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
# knit options
knitr::opts_chunk$set(echo = F,
                      results = 'markup',
                      fig.width = 4,
                      fig.height = 3,
                      fig.align = 'center',
                      message = F,
                      warning = F)

# packages
library(tidyverse)
library(faraway)
library(RSQLite)
library(skimr)
```

```{r, echo=FALSE}
ball <- dbConnect(drv=RSQLite::SQLite(), dbname="../../data/database.sqlite")
```

```{r, echo =FALSE}
batting <- dbGetQuery(ball, "
                            SELECT
                            sum(ab) AS AT_BAT,
                            player_id,
                            sum(r) AS RUNS,
                            sum(hr) AS HOME_RUNS, 
                            sum(triple) AS TRIPLE,
                            sum(double) AS DOUBLE,
                            (sum(h) -  sum(hr) - sum(triple) - sum(double)) AS SINGLES,
                            sum(bb) AS WALKS,
                            sum(ibb) AS INT_WALKS,
                            sum(sb) AS STOLEN_BASES,
                            sum(hbp) AS HIT_BY_PITCH
                            
                            FROM
                            batting
                            where
                            year > 2000
                            group by
                            player_id
                            
                   ")
sumbat <- batting
batting <- subset(batting,batting[,1]>100)


for (x in 3:11) 
{
  batting[,x] <- batting[,x] / 1
}

set.seed(5)
batting <- sample_n(batting, 500, replace = FALSE)

players <- dbGetQuery(ball, "
                      
                      SELECT 
                      player_id,
                      (weight / POWER(height, 2)) *703 AS BMI,
                      bats as HAND
                      FROM
                      player
                      
                      ")




players <- subset(players, players[,2]>0)



for ( x in 1:17918)
{
  if(players[x,2] <= 18.5) {players[x,2] <- "U"} 
  else if(players[x,2] <= 24.9) {players[x,2] <- "H"}
  else if(players[x,2] <= 29.9) {players[x,2] <- "O"}
  else {players[x,2] <- "B"}
}

batting <-  merge(batting, players, by="player_id" )
```

# Sam

```{r}
summary(lm(RUNS~HOME_RUNS + SINGLES + WALKS + STOLEN_BASES, data = batting))
```

# Introduction

For our project, we are looking at the "History of Baseball" data set, which is a record of all the stats of baseball players who have played in the MLB up until the year 2015. We are interested in seeing which of our predictors (singles, doubles, triples, home runs, walks, intentional walks, hit by pitches, stolen bases, player BMIs, and batting hand) contribute the most to runs scored by individual players. So far, we have seen that our sample aligns with overall MLB stats (eg. the batting average for our sample is 0.246 compared to the MLB’s 0.250) and all of our model assumptions hold, which allowed us to construct a simple linear model with doubles as our predictor. Now, we will be selecting predictors to construct potential multiple linear models and selecting the best one.

# Coefficient Interpretations

$\beta_1$: Every additional home run a player hits is associated with an increase of about 0.977261 mean runs, after accounting for singles, walks and stolen bases.

$\beta_2$: Every additional single a player hits is associated with an increase of about 0.400909 mean runs, after accounting for home runs, walks and stolen bases.

$\beta_3$: Every additional walk a player earns is associated with an increase of about 0.268561 mean runs, after accounting for home runs, singles and stolen bases.

$\beta_4$: Every additional stolen base a player earns is associated with an increase of about 0.268561 mean runs, after accounting for home runs, singles and walks.

**Significance Tests**

Looking at the p-values for each of our coefficients, we can see that all of our coefficients are significant at the 0.05 level.
