---
title: "126 Data Project, Step 4"
date: "Sam Ream, Valeria Lopez, Skyler Yee"
output:
  pdf_document:
    latex_engine: xelatex
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
# knit options
knitr::opts_chunk$set(echo = F,
                      results = 'markup',
                      fig.width = 4,
                      fig.height = 3,
                      fig.align = 'center',
                      message = F,
                      warning = F)

# packages
library(tidyverse)
library(faraway)
library(RSQLite)
library(skimr)
library(GGally)
library(tidymodels)
library(leaps)
library(glmnet)
```

```{r, echo=FALSE}
ball <- dbConnect(drv=RSQLite::SQLite(), dbname="../../data/database.sqlite")
```

```{r, echo =FALSE}
batting <- dbGetQuery(ball, "
                            SELECT
                            sum(ab) AS AT_BAT,
                            player_id,
                            sum(r) AS RUNS,
                            sum(hr) AS HOME_RUNS, 
                            sum(triple) AS TRIPLE,
                            sum(double) AS DOUBLE,
                            (sum(h) -  sum(hr) - sum(triple) - sum(double)) AS SINGLES,
                            sum(bb) AS WALKS,
                            sum(ibb) AS INT_WALKS,
                            sum(sb) AS STOLEN_BASES,
                            sum(hbp) AS HIT_BY_PITCH
                            
                            FROM
                            batting
                            where
                            year > 2000
                            group by
                            player_id
                            
                   ")
sumbat <- batting
batting <- subset(batting,batting[,1]>100)


for (x in 3:11) 
{
  batting[,x] <- batting[,x] / 1
}

set.seed(5)
batting <- sample_n(batting, 500, replace = FALSE)

players <- dbGetQuery(ball, "
                      
                      SELECT 
                      player_id,
                      (weight / POWER(height, 2)) *703 AS BMI,
                      bats as HAND
                      FROM
                      player
                      
                      ")




players <- subset(players, players[,2]>0)



for ( x in 1:17918)
{
  if(players[x,2] <= 18.5) {players[x,2] <- "U"} 
  else if(players[x,2] <= 24.9) {players[x,2] <- "H"}
  else if(players[x,2] <= 29.9) {players[x,2] <- "O"}
  else {players[x,2] <- "B"}
}

batting <-  merge(batting, players, by="player_id" )
```

# Summary

# Step 1

# Step 2

```{r fig.cap= "Scatter plot of the relationship between Runs and Doubles"}
par(mfrow = c(1, 2))
plot(batting$DOUBLE, batting$RUNS, ylab = "Runs", xlab = "Doubles")
abline(0,2.4, col="red",)
abline(-150,2.4, col="blue",)
abline(200,2.4, col="blue",)
```

```{r fig.cap= "QQ-Plot showing that the data is not normally distributed"}
qqnorm(batting$RUNS)
qqline(batting$RUNS)
```

**- Linearity:** All the points on the relationship plot above are arranged in a very linear way without transformations

**- Constant Variance:** Almost all of the points have a similar distance from a proposed straight line.

**- Independence:** With the knowledge that one batter hitting the ball well enough to get a double does not affect the likelihood of the next batter doing the same, we know that the predictors are independent of one another.

**- Normality:** While our errors do not appear to be normally distributed, our large sample size allows us to leverage the Central Limit Theorem to make meaningful analysis.

```{r include=FALSE}

slm <- summary(g)
anova_lm <- anova(g)
```

# Hypothesis Testing

## Significance Test

$$H_0: \beta_i=0$$

$$H_a: \beta_1 \neq 0$$

$$\alpha = 0.05$$

```{r include=FALSE}
test_stat <- slm$coefficients[,3][2]
test_stat #102.942 # from slm
p_value <- slm$coefficients[,4][2]
p_value
```

$\textbf{Test Statistic =}\ 102.942$

$\textbf{P Value} \approx 0$

We reject $H_0$ at 0.05 level. Thus, the amount of doubles a player hits is a significant predictor of how many runs the player scores.

# Fit of Model

The $R^2$ value of our model is 0.9551, which means that the model explains 95.51% of the variance of the recorded events. Additionally, the residual plot in figure 3 shows how the data points share a similar spread which implies that the model is a good fit.

```{r include=FALSE}
summary(lm(batting$DOUBLE ~ batting$RUNS))
```

```{r fig.cap= "Residual Plot of the fitted model"}
model <- lm(batting$RUNS ~ batting$DOUBLE)

res <- resid(model)

plot(fitted(model), res, ylab = "Residuals", xlab = "Fitted Model")
```

# Computational Models

For our computational models, we used the predictors: Total Intentional Walks, Singles, Triples, Stolen Bases, and Home Runs obtained in a career. We selected these predictors because of their low correlation in addition to their interesting relation to obtained Runs.

### Model 1 - Full Model ( $\Omega$ )

$\mathbb{E}[Y]=\text{Intercept } + \text{Intentional Walks }+\text{Singles }+\text{Triples }+\text{Stolen Bases }+\text{Home Runs}+\epsilon$

### Model 2 - Reduced Model ( $\omega$ )

$\mathbb{E}[Y]=\text{Intercept } + \text{Singles }+\text{Triples }+\text{Home Runs}+\epsilon$

### Comparison:

$H_0: \beta \in \omega :\text{The Reduced Model is sufficient}$

$H_\alpha: \beta \in \Omega \text{\\} \omega \in w :\text{The Reduced Model is not sufficient}$

```{r}
model_1 <- lm(RUNS~INT_WALKS + SINGLES + TRIPLE + STOLEN_BASES + HOME_RUNS, data = batting)
model_2 <- lm(RUNS~SINGLES + TRIPLE + HOME_RUNS, data = batting)
anova(model_1, model_2)
```

### Conclusion

As we rejected \$H_0\$ in favor for \$H\_\\alpha\$, we can determine that the reduced model does not model the data well enough to justify the reduction in predictors. As such, we decided to use model 1, the full model, as our computational model.

# Statistical Model

We used a stepwise search to create the best model for our data. For a size of 4 predictors the variables home runs, singles, walks, and stolen bases create a well fit model.

# Final Model Selection

Between the two models we created, the statistical model and computational model, we selected the statistical model. The reason behind this selection is that the statistical model has a larger $R_{adj}^2$ value and we want to explain as much of the variance as possible in our model.

# Step 3

# Step 4

We implemented a Ridge and Lasso regression on the entire set of predictors to make a model that fits well. While the Lasso regression did not eliminate any predictors, it did result in the best fit of all the previoud models as can be demonstrated in the plot and table below.






```{r}

y <- batting$RUNS
x <- data.matrix(batting[, c('TRIPLE', 'HOME_RUNS', 'SINGLES', 'WALKS', 'DOUBLE', 'INT_WALKS', 'STOLEN_BASES', 'HIT_BY_PITCH')])

stat_model <- lm(RUNS~HOME_RUNS + SINGLES + WALKS + STOLEN_BASES, data = batting)
slm <- lm(RUNS~DOUBLE, data = batting)


# find optimal lambda value that minimizes test MSE
best_lambda_r <- cv_model_ridge$lambda.min

cv_model <- cv.glmnet(x, y, alpha = 1)

# find optimal lambda value that minimizes test MSE
best_lambda <- cv_model$lambda.min

model <- glmnet(x, y, alpha = 1)

ridge_model <- glmnet(x, y, alpha = 0)

cv_model_ridge <- cv.glmnet(x, y, alpha = 0)

```





```{r, echo =FALSE}
batting <- dbGetQuery(ball, "
                            SELECT
                            sum(ab) AS AT_BAT,
                            player_id,
                            sum(r) AS RUNS,
                            sum(hr) AS HOME_RUNS, 
                            sum(triple) AS TRIPLE,
                            sum(double) AS DOUBLE,
                            (sum(h) -  sum(hr) - sum(triple) - sum(double)) AS SINGLES,
                            sum(bb) AS WALKS,
                            sum(ibb) AS INT_WALKS,
                            sum(sb) AS STOLEN_BASES,
                            sum(hbp) AS HIT_BY_PITCH
                            
                            FROM
                            batting
                            where
                            year > 2000
                            group by
                            player_id
                            
                   ")
sumbat <- batting
batting <- subset(batting,batting[,1]>100)

set.seed(4)
batting <- sample_n(batting, 500, replace = FALSE)

players <- dbGetQuery(ball, "
                      
                      SELECT 
                      player_id,
                      (weight / POWER(height, 2)) *703 AS BMI,
                      bats as HAND
                      FROM
                      player
                      
                      ")




players <- subset(players, players[,2]>0)



for ( x in 1:17918)
{
  if(players[x,2] <= 18.5) {players[x,2] <- "U"} 
  else if(players[x,2] <= 24.9) {players[x,2] <- "H"}
  else if(players[x,2] <= 29.9) {players[x,2] <- "O"}
  else {players[x,2] <- "B"}
}

batting <-  merge(batting, players, by="player_id" )

```
```{r}
x <- data.matrix(batting[, c('TRIPLE', 'HOME_RUNS', 'SINGLES', 'WALKS', 'DOUBLE', 'INT_WALKS', 'STOLEN_BASES', 'HIT_BY_PITCH')])









```

```{r}
ridge_model_y_predictions <- data.frame(Prediction = 1:500, Actual= 1:500)
ridge_model_y_predictions[,1] <- predict(ridge_model, s = best_lambda_r, newx = x)
ridge_model_y_predictions[,2] <- batting$RUNS

lasso_model_y_predictions <- data.frame(Prediction = 1:500, Actual = 1:500)
lasso_model_y_predictions[,1] <- predict(cv_model, s = best_lambda, newx = x)
lasso_model_y_predictions[,2] <- batting$RUNS


stat_model_y_predictions <- data.frame(Prediction = 1:500, Actual = 1:500)
stat_model_y_predictions[,1] <- predict(stat_model, newdata = batting)
stat_model_y_predictions[,2] <- batting$RUNS

slm_y_predictions <- data.frame(Prediction = 1:500, Actual = 1:500)
slm_y_predictions[,1] <- predict(slm, newdata = batting)
slm_y_predictions[,2] <- batting$RUNS
```

```{r, fig.width=7,fig.height=2, fig.cap="A comparison of the Linear Models"}
ggplot() + 
  
  geom_point(aes(x=Prediction, y=Actual, color="SLM"), slm_y_predictions, size =0.25 )+
  geom_smooth(aes(x=Prediction, y=Actual, color="SLM"), slm_y_predictions, method = lm, fullrange =TRUE, size = 0.75)+
  
  geom_point(aes(x=Prediction, y=Actual, color="MLR"), stat_model_y_predictions,size = 0.25) + 
  geom_smooth(aes(x=Prediction, y=Actual, color="MLR"), stat_model_y_predictions, method = lm, fullrange =TRUE, size = 0.75)+
  
  geom_point(aes(x=Prediction, y=Actual, color="Ridge Model"), ridge_model_y_predictions, size =0.25)+
  geom_smooth(aes(x=Prediction, y=Actual, color="Ridge Model"), ridge_model_y_predictions, method = lm, fullrange =TRUE, size = 0.75)+
  
  geom_point(aes(x=Prediction, y=Actual, color="Lasso Model"), lasso_model_y_predictions, size =0.25 )+
  geom_smooth(aes(x=Prediction, y=Actual, color="Lasso Model"), lasso_model_y_predictions, method = lm, fullrange =TRUE, size = 0.75)+
  
  labs(color = "Model")
```
