---
title: "126 Data Project, Step 2"
date: "Sam Ream, Valeria Lopez, Skyler Yee"
output:
  pdf_document:
    latex_engine: xelatex
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
# knit options
knitr::opts_chunk$set(echo = F,
                      results = 'markup',
                      fig.width = 4,
                      fig.height = 3,
                      fig.align = 'center',
                      message = F,
                      warning = F)

# packages
library(tidyverse)
library(faraway)
library(RSQLite)
library(skimr)
library(GGally)
library(tidymodels)
library(leaps)
```

```{r, echo=FALSE}
ball <- dbConnect(drv=RSQLite::SQLite(), dbname="../../data/database.sqlite")
```

```{r, echo =FALSE}
batting <- dbGetQuery(ball, "
                            SELECT
                            sum(ab) AS AT_BAT,
                            player_id,
                            sum(r) AS RUNS,
                            sum(hr) AS HOME_RUNS, 
                            sum(triple) AS TRIPLE,
                            sum(double) AS DOUBLE,
                            (sum(h) -  sum(hr) - sum(triple) - sum(double)) AS SINGLES,
                            sum(bb) AS WALKS,
                            sum(ibb) AS INT_WALKS,
                            sum(sb) AS STOLEN_BASES,
                            sum(hbp) AS HIT_BY_PITCH
                            
                            FROM
                            batting
                            where
                            year > 2000
                            group by
                            player_id
                            
                   ")
sumbat <- batting
batting <- subset(batting,batting[,1]>100)


for (x in 3:11) 
{
  batting[,x] <- batting[,x] / 1
}

set.seed(5)
batting <- sample_n(batting, 500, replace = FALSE)

players <- dbGetQuery(ball, "
                      
                      SELECT 
                      player_id,
                      (weight / POWER(height, 2)) *703 AS BMI,
                      bats as HAND
                      FROM
                      player
                      
                      ")




players <- subset(players, players[,2]>0)



for ( x in 1:17918)
{
  if(players[x,2] <= 18.5) {players[x,2] <- "U"} 
  else if(players[x,2] <= 24.9) {players[x,2] <- "H"}
  else if(players[x,2] <= 29.9) {players[x,2] <- "O"}
  else {players[x,2] <- "B"}
}

batting <-  merge(batting, players, by="player_id" )
```

# Sam

# Valeria

```{r}
lm(RUNS~INT_WALKS + SINGLES + TRIPLE + STOLEN_BASES + HOME_RUNS, data = batting) %>% tidy()
```

```{r}
lm(RUNS~SINGLES + TRIPLE + HOME_RUNS, data = batting) %>% tidy()
```

```{r}
stat_model <- regsubsets(RUNS~HOME_RUNS + TRIPLE + DOUBLE + SINGLES + WALKS + INT_WALKS + STOLEN_BASES + HIT_BY_PITCH, data = batting,
                  method = 'seqrep',
                  nbest = 1,
                  nvmax = 4)
summary(stat_model)
```

We used a stepwise search to create the best model for our data. For a size of 4 predictors the variables home runs, singles, walks, and stolen bases create a well fit model. Our adjusted R^2 is .9925. 


## Stat model residual plot

```{r}
stat_model <- lm(RUNS~HOME_RUNS + SINGLES + WALKS + STOLEN_BASES, data = batting)
res <- resid(stat_model)
plot(fitted(stat_model), res, ylab = 'Residuals', xlab = 'Fitted Model')
abline(0,0)
```
```{r}
summary(lm(RUNS~HOME_RUNS + SINGLES + WALKS + STOLEN_BASES, data = batting))
```


## CI 
```{r}
chosen_model <- lm(RUNS~ HOME_RUNS + SINGLES + WALKS + STOLEN_BASES, data = batting)
```

```{r}
confint(chosen_model, level = 0.95)
```

```{r}
x_bar <- batting %>% select(HOME_RUNS, SINGLES, WALKS, STOLEN_BASES) %>% summarize(across(everything(), mean))
x_bar
```

```{r}
predict(chosen_model, newdata = x_bar, interval = 'confidence', level = 0.95)
```
With 95% confidence, the mean predicted value is estimated to be between 204.71 and 208.50.



## PI
```{r}
comb_x <- data.frame(batting$HOME_RUNS[3], batting$SINGLES[5], batting$WALKS[10], batting$STOLEN_BASES[7])
colnames(comb_x) <- c("HOME_RUNS", "SINGLES", "WALKS", "STOLEN_BASES")
comb_x

batting2 <- batting %>% slice(-as.numeric(unlist(comb_x)))
```
```{r}
predict(chosen_model, newdata = comb_x , interval = 'prediction', level = 0.95)
```

With 95% confidence, the predicted value for a combination of 2 home runs, 5 singles, 10 walks, and 7 stolen bases is between 188.06 and 273.59. 



## Summary
We sampled our data randomly from years 2000-2015 to get an accurate representation of the population and represent the changing baseball strategy. We analyzed multiple quantitative variables along with BMI and the batting hand of players in relation to runs. While checking the assumptions for linear regression, we found that our data was not normally distributed but were able to continue with our analysis due to the central limit theorem. Through hypothesis testing, we found that doubles is a significant predictor of how many runs the player scores and the residual plot showed that our model was a good fit. In order to create computational models, we chose variables that had low correlation, which consisted of intentional walks singles, triples, stole bases, and home runs. Our other computation model was a reduced version. To create a statistical model, we used stepwise search with four predictors, resulting in a model with home runs, singles, walks, and stolen bases. Then we tested the fit of the model and created confidence and prediction intervals.          





# Skyler
