---
title: "126 Data Project, Step 4"
date: "Sam Ream, Valeria Lopez, Skyler Yee"
output:
  pdf_document:
    latex_engine: xelatex
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
# knit options
knitr::opts_chunk$set(echo = F,
                      results = 'markup',
                      fig.width = 4,
                      fig.height = 3,
                      fig.align = 'center',
                      message = F,
                      warning = F)

# packages
library(tidyverse)
library(faraway)
library(RSQLite)
library(skimr)
library(GGally)
library(tidymodels)
library(leaps)
library(glmnet)
library(gridExtra)
```
```{r, echo=FALSE}
ball <- dbConnect(drv=RSQLite::SQLite(), dbname="../../data/database.sqlite")
```
```{r, echo =FALSE}
batting <- dbGetQuery(ball, "
                            SELECT
                            sum(ab) AS AT_BAT,
                            player_id,
                            sum(r) AS RUNS,
                            sum(hr) AS HOME_RUNS, 
                            sum(triple) AS TRIPLE,
                            sum(double) AS DOUBLE,
                            (sum(h) -  sum(hr) - sum(triple) - sum(double)) AS SINGLES,
                            sum(bb) AS WALKS,
                            sum(ibb) AS INT_WALKS,
                            sum(sb) AS STOLEN_BASES,
                            sum(hbp) AS HIT_BY_PITCH
                            
                            FROM
                            batting
                            where
                            year > 2000
                            group by
                            player_id
                            
                   ")
sumbat <- batting
batting <- subset(batting,batting[,1]>100)


for (x in 3:11) 
{
  batting[,x] <- batting[,x] / 1
}

set.seed(5)
batting <- sample_n(batting, 500, replace = FALSE)

players <- dbGetQuery(ball, "
                      
                      SELECT 
                      player_id,
                      (weight / POWER(height, 2)) *703 AS BMI,
                      bats as HAND
                      FROM
                      player
                      
                      ")




players <- subset(players, players[,2]>0)



for ( x in 1:17918)
{
  if(players[x,2] <= 18.5) {players[x,2] <- "U"} 
  else if(players[x,2] <= 24.9) {players[x,2] <- "H"}
  else if(players[x,2] <= 29.9) {players[x,2] <- "O"}
  else {players[x,2] <- "B"}
}

batting <-  merge(batting, players, by="player_id" )
```


# Introduction

Using the "History of Baseball" data set, we analyzed how our predictors (singles, doubles, triples, home runs, walks, intentional walks, hit by pitches, stolen bases, BMI, and batting hand) affected the runs scores by individual players. We sampled player statistics randomly from games played between 2000-2015, which allowed us to get an accurate representation of the population of all players who played between 2000 and 2015. Using both Ridge Regression and LASSO, we shrunk the size of some predictors to obtain estimates with smaller variance for higher precision. 

# Summary of previous steps

## Step 1



## Step 2



## Step 3



# Ridge Regression


# Lasso Regression


# Comparison of the Models


# Investigation - Principle component analysis


# Conclusion

